---
title: "Stickleback case study analysis"
author: "Max Czapanskiy"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "../figures/",
  dpi = 600
)
library(cowplot)
library(glue)
library(lubridate)
library(tidyverse)
```

```{r load_results}
breath_cv_id <- "202208021602"
breath_cv_dir <- file.path("analysis/data/derived_data/cvtrials", breath_cv_id)
breath_results <- read_csv(file.path(breath_cv_dir, "_results.csv"),
                           show_col_types = FALSE)

lunge_cv_id <- "202208021603"
lunge_cv_dir <- file.path("analysis/data/derived_data/cvtrials", lunge_cv_id)
lunge_results <- read_csv(file.path(lunge_cv_dir, "_results.csv"),
                           show_col_types = FALSE)

results <- rbind(
  mutate(breath_results, behavior = "Breath"),
  mutate(lunge_results, behavior = "Lunge")
)

results_long <- results %>%
  pivot_longer(-c(behavior, trial, rf_winsz), names_to = "metric") %>%
  mutate(model = factor(str_sub(metric, 1, 2),
                        levels = c("sb", "rf"),
                        labels = c("Stickleback", "Random forest")),
         metric = factor(str_sub(metric, 4),
                         levels = c("f1", "delta_r"),
                         labels = c("F1", "delta[r]"))) %>%
  relocate(model, .after = trial)
```

```{r load_raw}
events <- arrow::read_parquet("analysis/data/raw_data/events.parquet")
sensors <- arrow::read_parquet("analysis/data/raw_data/sensors.parquet")

# Fix an error where 4 deployments have missing speed values
missing_speed <- c("bw180828-49", "bw180830-42", "bw180830-48", "bw180905-42")
events <- filter(events, !deployid %in% missing_speed)
sensors <- filter(sensors, !deployid %in% missing_speed)
```

## Case study data

```{r}
duration_hr <- function(dt) {
  as.numeric(last(dt) - first(dt), unit = "hours")
}
details <- function(median, min, max) {
  glue("({signif(median, 3)}; {signif(min, 3)} - {signif(max, 3)})")
}
sensors %>% 
  group_by(deployid) %>% 
  summarize(dur_hr = duration_hr(datetime)) %>% 
  left_join(
    events %>% 
      group_by(deployid, event) %>% 
      summarize(N = n(), .groups = "drop"),
    by = "deployid"
  ) %>% 
  group_by(event) %>% 
  summarize(across(c(dur_hr, N), 
                   list(deployments = length,
                        sum = sum, 
                        median = median, 
                        min = min, 
                        max = max))) %>% 
  transmute(Species = "Blue whale",
            Deployments = N_deployments,
            Behavior = str_to_title(event),
            `Duration [hr]` = round(dur_hr_sum),
            `Duration [hr]2` = details(dur_hr_median, dur_hr_min, dur_hr_max),
            N = N_sum,
            N2 = details(N_median, N_min, N_max))
```


## Model performance metrics

- 2x models (Stickleback, random forest)
- 2x behaviors (breathing, lunge feeding)
- 2x performance metrics ($F1$, $\delta_r$)

As figure:

```{r model_perf}
f1_deltar_plots <- results_long %>%
  group_by(metric, behavior) %>%
  group_map(function(results, group) {
    label_fn <- if (group$metric == "delta[r]") {
      scales::percent
    } else {
      scales::number
    }
    y_title <- parse(text = as.character(group$metric))

    ggplot(results, aes(model, value)) +
      geom_boxplot() +
      scale_y_continuous(y_title, labels = label_fn) +
      expand_limits(y = 0) +
      labs(x = group$behavior) +
      theme_bw(base_size = 10) +
      theme()
  })

f1_plot <- filter(results_long, metric == "F1") %>% 
  ggplot(aes(model, value, color = model)) +
  geom_boxplot() +
  scale_y_continuous("F1") +
  scale_color_manual(values = c("Random forest" = "red",
                                "Stickleback" = "blue")) +
  expand_limits(y = c(0, 1)) +
  facet_grid(cols = vars(behavior)) +
  theme_bw() +
  theme(axis.title.x = element_blank(),
        legend.position = "none",
        strip.background = element_rect(fill = "white"))

deltar_plot <- filter(results_long, metric == "delta[r]") %>% 
  ggplot(aes(model, value, color = model)) +
  geom_boxplot() +
  scale_y_continuous(bquote(delta[r]), labels = scales::percent) +
  scale_color_manual(values = c("Random forest" = "red",
                                "Stickleback" = "blue")) +
  facet_grid(cols = vars(behavior)) +
  theme_bw() +
  theme(axis.title.x = element_blank(),
        legend.position = "none",
        strip.background = element_rect(fill = "white"))


f1_deltar_plot <- plot_grid(f1_plot, deltar_plot,
                            nrow = 2,
                            align = "v",
                            axis = "lr",
                            labels = "auto")
```

```{r model_perf_plot, fig.dim=c(4.72, 4.72)}
f1_deltar_plot
```

As table:

```{r model_perf_table, comment=""}
format_metric <- function(metric, metric_mean, metric_sd) {
  format_string <- ifelse(
    metric == "delta[r]",
    "$%.3g\\%% \\pm %.3g\\%%$",
    "$%.3g \\pm %.3g$"
  )
  replace_exp <- function(s) {
    e <- parse_number(str_match(s, "([0-9.]+e[+-][0-9]+)")[,2])
    ifelse(is.na(e),
           s,
           str_replace(s, "[0-9.]+e[+-][0-9]+", as.character(signif(e, 3))))
  }
  sprintf(format_string, metric_mean, metric_sd) %>%
    replace_exp()
}

results_long %>%
  group_by(metric, model, behavior) %>%
  summarize(across(-c(trial, rf_winsz), list(mean = mean, sd = sd)),
            .groups = "drop") %>%
  mutate(
    across(starts_with("value"),
           ~ ifelse(metric == "delta[r]", .x * 100, .x)),
    label = format_metric(metric, value_mean, value_sd),
    metric = sprintf("$%s$", ifelse(metric == "F1", "F1", "\\delta_r"))
  ) %>%
  select(-value_mean, -value_sd) %>%
  pivot_wider(names_from = c(behavior, model),
              names_sep = " ",
              values_from = label) %>%
  select(metric, starts_with("Breath"), starts_with("Lunge")) %>%
  knitr::kable(escape = FALSE,
               col.names = c("", rep(c("Stickleback", "Random forest"), 2)),
               booktabs = TRUE) %>%
  kableExtra::add_header_above(c("", Breath = 2, Lunge = 2))
```

## Prediction outcomes

Stickleback predicted mostly TPs but random forest mostly FPs, which inflated the $\delta_r$.

```{r pred_outcomes}
read_predictions <- function(cv_dir) {
  cv_behavior <- substr(readLines(file.path(cv_dir, "params.txt"))[1], 12, 100)
  read_trial <- function(trial_dir) {
    model_predictions <- dir(trial_dir, 
                             pattern = "predictions.rds", 
                             full.names = TRUE)
    read_model <- function(pred_path) {
      readRDS(pred_path) %>% 
        mutate(trial = basename(trial_dir),
               model = if (substr(basename(pred_path), 2, 3) == "rf") {
                 "Random forest"
               } else {
                 "Stickleback"
               })
    }
    map_dfr(model_predictions, read_model)
  }
  
  map_dfr(dir(cv_dir, pattern = "[0-9]{2}", full.names = TRUE), read_trial) %>% 
    mutate(behavior = cv_behavior)
}

predictions_long <- rbind(
  read_predictions(breath_cv_dir),
  read_predictions(lunge_cv_dir)
)

predictions_summary <- predictions_long %>% 
  group_by(behavior, model, trial, outcome) %>% 
  summarize(N = n(), .groups = "drop_last") %>% 
  mutate(frac = N / sum(N)) %>% 
  ungroup() %>% 
  mutate(behavior = str_to_title(behavior),
         outcome = factor(outcome, levels = c("TP", "FP", "FN")),
         model = factor(model, levels = c("Stickleback", "Random forest")))

outcome_plot <- ggplot(predictions_summary, aes(outcome, frac, color = model)) +
  geom_boxplot(size = 0.5) +
  expand_limits(y = c(0, 1)) +
  scale_y_continuous("Predictions", labels = scales::percent) +
  scale_color_manual(values = c("Stickleback" = "blue",
                                "Random forest" = "red")) +
  facet_grid(cols = vars(behavior), scales = "free_y") +
  theme_bw() +
  theme(axis.title.x = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        strip.background = element_rect(fill = NA))
```

```{r outcome_plot, fig.dim=c(4.72, 4.72)}
outcome_plot
```

## Window sizes

Optimal window size for the random forest.

```{r rf_win}
read_winsz <- function(cv_dir) {
  cv_behavior <- substr(readLines(file.path(cv_dir, "params.txt"))[1], 12, 100)
  file.path(cv_dir, "_results.csv") %>% 
    read_csv(show_col_types = FALSE) %>% 
    transmute(trial, winsz = rf_winsz, behavior = cv_behavior)
}

rbind(
  read_winsz(breath_cv_dir),
  read_winsz(lunge_cv_dir)
) %>% 
  group_by(behavior) %>% 
  summarize(across(winsz, list(mean = mean, sd = sd))) %>% 
  mutate(across(starts_with("winsz"), ~ round(.x / 10, 1)))
```

## False positives in random forest

Random forest false positives were clustered around true positives.

```{r rf_fp}
dive <- as.POSIXct(c("2018-09-05 11:00", "2018-09-05 11:15"))
lunge_sensors <- readRDS(file.path(lunge_cv_dir, "01", "test_sensors.rds")) %>% 
  filter(deployid == "bw180905-49",
         between(datetime, dive[1], dive[2]))
lunge_actual <- readRDS(file.path(lunge_cv_dir, "01", "test_events.rds")) %>% 
  filter(deployid == lunge_sensors$deployid[1],
         between(datetime, 
                 min(lunge_sensors$datetime), 
                 max(lunge_sensors$datetime))) %>% 
  left_join(lunge_sensors, by = c("deployid", "datetime"))
lunge_rf <- file.path(lunge_cv_dir, "01", "_rfpredictions.rds") %>% 
  readRDS() %>% 
  filter(deployid == lunge_sensors$deployid[1],
         between(datetime, 
                 min(lunge_sensors$datetime), 
                 max(lunge_sensors$datetime))) %>% 
  left_join(lunge_sensors, by = c("deployid", "datetime")) %>%
  mutate(outcome = factor(outcome, levels = c("FP", "FN", "TP"))) %>% 
  arrange(outcome)
lunge_sb <- file.path(lunge_cv_dir, "01", "_sbpredictions.rds") %>% 
  readRDS() %>% 
  filter(deployid == lunge_sensors$deployid[1],
         between(datetime, 
                 min(lunge_sensors$datetime), 
                 max(lunge_sensors$datetime))) %>% 
  left_join(lunge_sensors, by = c("deployid", "datetime")) %>%
  mutate(outcome = factor(outcome, levels = c("FP", "FN", "TP"))) %>% 
  arrange(outcome)

inset <- as.POSIXct(c("2018-09-05 11:04:00", "2018-09-05 11:04:10"))
inset_sensors <- filter(lunge_sensors, between(datetime, inset[1], inset[2]))
inset_actual <- filter(lunge_actual, between(datetime, inset[1], inset[2]))
inset_rf <- filter(lunge_rf, between(datetime, inset[1], inset[2]))
inset_sb <- filter(lunge_sb, between(datetime, inset[1], inset[2]))

outcome_colors <- c(TP = "#56B4E9", FP = "#009E73", FN = "#E69F00")
outcome_shapes <- c(TP = 21, FP = 22, FN = 22)

dive_plot <- ggplot(lunge_sensors, aes(datetime, depth)) +
  geom_line(size = 0.5, color = "grey50") +
  geom_point(aes(fill = outcome, shape = outcome), 
             lunge_rf, 
             color = "#00000088") +
  geom_point(aes(fill = outcome), 
             lunge_sb, 
             shape = 24, color = "#00000088", fill = outcome_colors["TP"],
             position = position_nudge(y = -7.5)) +
  geom_point(data = lunge_actual, 
             shape = 25, 
             position = position_nudge(y = 7.5)) +
  annotate("rect", 
           xmin = inset[1] - 5, 
           xmax = inset[2] + 5,
           ymin = min(inset_sensors$depth) - 7.5,
           ymax = max(inset_sensors$depth) + 7.5,
           fill = NA, color = "magenta", size = 0.5) +
  scale_x_datetime(date_labels = "%b-%d %H:%M") +
  scale_y_reverse("Depth (m)") +
  scale_fill_manual(values = outcome_colors) +
  scale_shape_manual(values = outcome_shapes) +
  theme_bw() +
  theme(axis.title.x = element_blank(),
        legend.position = "none")

inset_plot <- ggplot(inset_sensors, aes(datetime, depth)) +
  geom_line(size = 0.5, color = "grey50") +
  geom_point(aes(fill = outcome, shape = outcome), 
             inset_rf, 
             color = "#00000088") +
  geom_point(data = inset_actual, 
             shape = 25, 
             position = position_nudge(y = 1)) +
  geom_point(aes(fill = outcome), 
             inset_sb, 
             shape = 24, color = "#00000088",
             position = position_nudge(y = -1)) +
  expand_limits(y = c(200, 220)) +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  scale_y_reverse("Depth (m)", breaks = c(200, 210, 220)) +
  scale_fill_manual(values = outcome_colors) +
  scale_shape_manual(values = outcome_shapes) +
  theme_bw(base_size = 9) +
  theme(axis.title.x = element_blank(),
        legend.position = "none",
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white", 
                                       color = "magenta", 
                                       size = 1))

pred_legend <- tibble(
  order = 5:1,
  shape = c(25, 24, outcome_shapes),
  color = c("black", rep("#00000088", 4)),
  fill = c("white", outcome_colors["TP"], outcome_colors),
  label = c("Actual", 
            "Stickleback TP", 
            "Random forest TP",
            "Random forest FP",
            "Random forest FN")
) %>% 
  ggplot(aes(x = 1, y = order, shape = shape, color = color, fill = fill)) +
  geom_point(size = 3) +
  geom_text(aes(label = label),
            color = "black",
            hjust = 0,
            size = 3,
            nudge_x = 0.25) +
  coord_cartesian(xlim = c(0.9, 2.5)) +
  scale_shape_identity() +
  scale_color_identity() +
  scale_fill_identity() +
  theme_void() +
  theme(plot.background = element_rect(fill = "white", color = NA))

ggdraw(dive_plot) +
  draw_plot(inset_plot, 
            x = 0.49, y = 0.7, 
            width = 0.5, height = 0.45,
            hjust = 0.5, vjust = 0.5) +
  draw_plot(pred_legend,
            x = 0.97, y = 0.1,
            width = 0.2, height = 0.4,
            hjust = 1, vjust = 0)
```

